---
title: "coffee-analysis.qmd"
format: html
editor: visual
---

## Setup

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(caret)
library(Metrics)
library(randomForest)
library(broom)
library(patchwork)

set.seed(42)
```

## Load data

```{r load}
coffee <- read_csv("df_arabica_clean.csv")  

glimpse(coffee)
summary(coffee$`Total Cup Points`)
```

# Filtering out the bad brews

Parse text-to-numbers (altitude), Collapse too-many categories (variety: 48→6, processing: 10→4), Fix spelling chaos (color), Drop missing-data rows, Remove outliers that would warp the regression line

```{r}
coffee <- coffee %>%
  mutate(
    Altitude_m = sapply(Altitude, function(x) {
      if (is.na(x)) return(NA)
      nums <- as.numeric(unlist(regmatches(x, gregexpr("[0-9]+", x))))
      mean(nums)
    })
  )

summary(coffee$Altitude_m)
coffee %>% select(`Country of Origin`, Altitude, Altitude_m) %>% slice(1:10)
```

```{r}
coffee <- coffee %>%
  mutate(Altitude_m = if_else(Altitude_m > 3000, Altitude_m / 3.281, Altitude_m))

summary(coffee$Altitude_m)
```

```{r}
coffee <- coffee %>%
  mutate(
    Variety_clean = case_when(
      Variety %in% c("Gesha", "Caturra", "Typica", "Bourbon", "Catuai") ~ Variety,
      is.na(Variety) | Variety == "unknown" ~ "Unknown",
      TRUE ~ "Other"
    )
  )

coffee %>% count(Variety_clean) %>% arrange(desc(n))
```

```{r}
coffee <- coffee %>%
  mutate(
    Processing_clean = case_when(
      str_detect(`Processing Method`, regex("washed|wet|lavado", ignore_case = TRUE)) ~ "Washed",
      str_detect(`Processing Method`, regex("honey|pulped|mossto", ignore_case = TRUE)) ~ "Honey",
      str_detect(`Processing Method`, regex("natural|dry", ignore_case = TRUE)) ~ "Natural",
      is.na(`Processing Method`) ~ NA_character_,
      TRUE ~ "Other"
    )
  )

coffee %>% count(Processing_clean)
```

```{r}
coffee <- coffee %>%
  mutate(
    Color_clean = case_when(
      is.na(Color) ~ NA_character_,
      str_detect(Color, regex("brow", ignore_case = TRUE)) ~ "brownish",
      str_detect(Color, regex("blue|bluish", ignore_case = TRUE)) ~ "blue-green",
      str_detect(Color, regex("pale", ignore_case = TRUE)) ~ "pale-yellow",
      str_detect(Color, regex("yell", ignore_case = TRUE)) & 
        str_detect(Color, regex("green", ignore_case = TRUE)) ~ "yellow-green",
      str_detect(Color, regex("yell", ignore_case = TRUE)) ~ "yellowish",
      str_detect(Color, regex("greenish", ignore_case = TRUE)) ~ "greenish",
      Color == "green" ~ "green",
      TRUE ~ "other"
    )
  )

coffee %>% count(Color_clean)
```

```{r}
coffee_clean <- coffee %>%
  filter(
    !is.na(Altitude_m),
    !is.na(Processing_clean),
    !is.na(Color_clean)
  )

cat("Rows after dropping missing:", nrow(coffee_clean), "\n")

# 2. z-score TCP, flag outliers (Camp's ±2 SD threshold)
coffee_clean <- coffee_clean %>%
  mutate(TCP_z = (`Total Cup Points` - mean(`Total Cup Points`)) / sd(`Total Cup Points`))

# how many outliers?
n_outliers <- sum(abs(coffee_clean$TCP_z) > 2)
cat("TCP outliers (|z| > 2):", n_outliers, 
    "(", round(100 * n_outliers / nrow(coffee_clean), 1), "%)\n")

# remove them
coffee_clean <- coffee_clean %>%
  filter(abs(TCP_z) <= 2)

cat("Final dataset:", nrow(coffee_clean), "rows\n")
```

```{r}
write_csv(coffee_clean, "df_coffee_cleaned.csv")
```

# EDA — looking at the beans

How do these beans stack up? (TCP histogram)

```{r}
ggplot(coffee_clean, aes(x = `Total Cup Points`)) +
  geom_histogram(bins = 20, fill = "tan", color = "white") +
  labs(title = "Distribution of Total Cup Points",
       x = "Total Cup Points",
       y = "Number of coffees") +
  theme_minimal()
```

## Reaching new heights (TCP vs altitude)

```{r}
ggplot(coffee_clean, aes(x = Altitude_m, y = `Total Cup Points`)) +
  geom_point(alpha = 0.5, color = "saddlebrown") +
  geom_smooth(method = "lm", color = "darkred", se = TRUE) +
  labs(title = "Does altitude predict cup quality?",
       x = "Altitude (meters)",
       y = "Total Cup Points") +
  theme_minimal()
```

## Beans around the world (TCP by country)

```{r}
top_countries <- coffee_clean %>% 
  count(`Country of Origin`) %>% 
  filter(n >= 5) %>% 
  pull(`Country of Origin`)

coffee_clean %>%
  filter(`Country of Origin` %in% top_countries) %>%
  ggplot(aes(x = reorder(`Country of Origin`, `Total Cup Points`, median),
             y = `Total Cup Points`)) +
  geom_boxplot(fill = "lightblue") +
  coord_flip() +
  labs(title = "Cup quality by country",
       subtitle = "(only countries with n ≥ 5 samples shown)",
       x = NULL,
       y = "Total Cup Points") +
  theme_minimal()
```

# Model 1 — Drawing the Line in the Beans

A multiple linear regression to test H1: do altitude, variety, processing, color, moisture, and defects predict Total Cup Points?

Note: country is NOT in this model — it goes into Model 2 as a random effect.

```{r}
m1 <- lm(`Total Cup Points` ~ Altitude_m + Variety_clean + Processing_clean +
                              Color_clean + `Moisture Percentage` +
                              `Category One Defects` + `Category Two Defects`,
         data = coffee_clean)

summary(m1)
```

# Model 2 — Country Roads, Bean Me Home

A mixed effects model that adds country as a random effect. Tests H2: does country explain variance beyond what other predictors capture?

```{r}
m2 <- lmer(`Total Cup Points` ~ Altitude_m + Variety_clean + Processing_clean +
                                Color_clean + `Moisture Percentage` +
                                `Category One Defects` + `Category Two Defects` +
                                (1 | `Country of Origin`),
           data = coffee_clean)

summary(m2)
```

```{r}
varcomp <- as.data.frame(VarCorr(m2))
icc <- varcomp$vcov[1] / sum(varcomp$vcov)
icc
```

```{r}
anova(m2, m1)
```

```{r}
AIC(m1, m2)
```

```{r}
coffee_clean <- coffee_clean %>% filter(Processing_clean != "Other")
cat("Rows after dropping single 'Other' processing case:", nrow(coffee_clean), "\n")
```

# Train/Test Split

Setting up 80/20 split for model comparison (Week 14 method).

```{r}
set.seed(42)

train_idx <- createDataPartition(coffee_clean$`Total Cup Points`, p = 0.8, list = FALSE)
train_data <- coffee_clean[train_idx, ]
test_data  <- coffee_clean[-train_idx, ]

cat("Training set:", nrow(train_data), "rows\n")
cat("Testing set:", nrow(test_data), "rows\n")
```

## Refitting M1 and M2 on training data

```{r}
# Refit M1 (lm) on training data
m1_train <- lm(`Total Cup Points` ~ Altitude_m + Variety_clean + Processing_clean +
                                    Color_clean + `Moisture Percentage` +
                                    `Category One Defects` + `Category Two Defects`,
               data = train_data)

# Refit M2 (lmer) on training data
m2_train <- lmer(`Total Cup Points` ~ Altitude_m + Variety_clean + Processing_clean +
                                      Color_clean + `Moisture Percentage` +
                                      `Category One Defects` + `Category Two Defects` +
                                      (1 | `Country of Origin`),
                 data = train_data)

# Predict on test data
m1_preds <- predict(m1_train, newdata = test_data)
m2_preds <- predict(m2_train, newdata = test_data, allow.new.levels = TRUE)

# Compute RMSE using Metrics::rmse (Camp's package from week 14)
m1_rmse <- rmse(test_data$`Total Cup Points`, m1_preds)
m2_rmse <- rmse(test_data$`Total Cup Points`, m2_preds)

cat("M1 (Linear Regression) test RMSE:", round(m1_rmse, 4), "\n")
cat("M2 (Mixed Effects)     test RMSE:", round(m2_rmse, 4), "\n")
```

# Model 3 — Can't See the Bean for the Trees

A random forest model. Tests H3: does a flexible tree-based model predict TCP more accurately than the linear or mixed-effects models?

```{r}
# Step 1: rename columns to have no spaces (randomForest hates spaces)
train_rf <- train_data %>%
  select(`Total Cup Points`, Altitude_m, Variety_clean, Processing_clean,
         Color_clean, `Moisture Percentage`, `Category One Defects`,
         `Category Two Defects`, `Country of Origin`) %>%
  rename(TCP = `Total Cup Points`,
         Moisture = `Moisture Percentage`,
         Defects_1 = `Category One Defects`,
         Defects_2 = `Category Two Defects`,
         Country = `Country of Origin`) %>%
  mutate(across(c(Variety_clean, Processing_clean, Color_clean, Country), as.factor))

test_rf <- test_data %>%
  select(`Total Cup Points`, Altitude_m, Variety_clean, Processing_clean,
         Color_clean, `Moisture Percentage`, `Category One Defects`,
         `Category Two Defects`, `Country of Origin`) %>%
  rename(TCP = `Total Cup Points`,
         Moisture = `Moisture Percentage`,
         Defects_1 = `Category One Defects`,
         Defects_2 = `Category Two Defects`,
         Country = `Country of Origin`) %>%
  mutate(across(c(Variety_clean, Processing_clean, Color_clean, Country), as.factor))

# Step 2: match factor levels between train and test
for (col in c("Variety_clean", "Processing_clean", "Color_clean", "Country")) {
  levels(test_rf[[col]]) <- levels(train_rf[[col]])
}

# Step 3: fit the random forest
set.seed(42)
rf_model <- randomForest(TCP ~ .,
                         data = train_rf,
                         ntree = 500,
                         importance = TRUE)

print(rf_model)
```

```{r}
# Predict on test set
rf_preds <- predict(rf_model, newdata = test_rf)

# Compute RMSE
rf_rmse <- rmse(test_rf$TCP, rf_preds)

cat("M3 (Random Forest) test RMSE:", round(rf_rmse, 4), "\n")
```

```{r}
varImpPlot(rf_model, main = "Variable Importance — Random Forest")
```

```{r}
# Final test RMSE comparison
results <- data.frame(
  Model = c("M1: Linear Regression", 
            "M2: Mixed Effects",
            "M3: Random Forest"),
  Test_RMSE = c(round(m1_rmse, 4), 
                round(m2_rmse, 4), 
                round(rf_rmse, 4))
) %>% 
  arrange(Test_RMSE)

results
```
