---
title: "Coding Categorical Variables"
author: 
  - "Student Name"
  - "DS303, SP25"
  - "Prof. Amber Camp"
date: 2/19/25
format: html
editor: visual
theme: spacelab
---

## Coding Categorical Predictors

Working with categorical predictors is an important aspect of building effective statistical models. Categorical predictors, also known as factors, represent qualitative data that can be organized into distinct groups, such as gender, geographic location, or product type.

In R, these predictors can be integrated into various modeling techniques, helping us identify relationships and make predictions based on different *levels* of these variables.

## UCLA Reference

There are many resources out there, but this is one I go back to again and again. I find it really easy to understand and to also use as a reference:

[**R LIBRARY CONTRAST CODING SYSTEMS FOR CATEGORICAL VARIABLES**](https://stats.oarc.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/)

## Set up your environment

### Load packages

```{r, echo = FALSE, message = FALSE}
library(lme4)
library(tidyverse)
```

### Load Data

Following the UCLA resource, we will use data frame called `hsb2` and focusing on the categorical variable `race`, which has four levels (1 = Hispanic, 2 = Asian, 3 = African American and 4 = Caucasian). We will use `write` as our dependent variable.

[Here](https://www.openintro.org/data/index.php?data=hsb2) is a brief description of `hsb2` data (but note that they have filled in some of their values a bit differently than we have below).

```{r}
# read in data from URL
hsb2 <- read.table('https://stats.idre.ucla.edu/stat/data/hsb2.csv', header=T, sep=",")

skim <- skimr::skim(hsb2)
summary(as.factor(hsb2$race))
```

Let's create `race.f`, which is the `race` data re-labeled and also turned into factors.

```{r}
hsb2$race.f <- factor(hsb2$race, labels=c("Hispanic", "Asian", "African-Am", "Caucasian"))

```

### Inspect your data

Inspect your data below using `skimr`, `summary`, and `ggplot2`. You should also check out [this resource](https://www.openintro.org/data/index.php?data=hsb2), which describes the dataset.

geom_jitter(width = 0.2, height = 0)

```{r}
skim <- skimr::skim(hsb2)
View(skim)

summary(hsb2)
summary(hsb2$race)
summary(hsb2$race.f)

ggplot(hsb2, aes(x = race.f, y = write)) +
  geom_boxplot()

ggplot(hsb2, aes(x = race.f, y = write)) +
  geom_jitter(width = 0.2, height = 0) 

```

## Coding Schemes

There are several different coding schemes, and you can also generate your own scheme for your own purposes. For our class, we will focus on **Dummy**, **Simple**, **Deviation**, and **Helmert** Coding.

We've already seen dummy coding, but it is good to see it again.

## Class activity

Split into four groups. Explore your assigned coding scheme and answer the below five questions. Then, apply your coding scheme to the `hsb2` data and walk us through its interpretation.

## Dummy Coding (a.k.a. Treatment Coding)

(Selena, Ashley, Lela)

1.  What comparisons are made with this coding scheme? How does this coding scheme affect the coefficients in regression models? (What do the coefficients represnt? What is the reference level?)\
    \

2.  When is this coding scheme preferred? What types of data or analysis scenarios make this coding method particularly useful?\
    \

3.  What are the advantages of this coding method? How does it simplify interpretation or enhance model performance?\
    \

4.  What are the limitations of this coding method? Are there situations where this method may lead to misleading results or interpretations?\
    \
    \

### Applying Dummy Coding

```{r}
# assign dummy coding
contrasts(hsb2$race.f)

# build and summarize your regression
dummy <- lm(write ~ race.f, data = hsb2)
summary(dummy)
```

Summarize these findings. What is your reference group, and how do the other groups compare with regards to their writing scores? What does the intercept value represent?\
\
(write here)\

## Simple Coding

(Cade, Brandon, Izack)

1.  What comparisons are made with this coding scheme? How does this coding scheme affect the coefficients in regression models? (What do the coefficients represnt? What is the reference level?)\

2.  When is this coding scheme preferred? What types of data or analysis scenarios make this coding method particularly useful?\

3.  What are the advantages of this coding method? How does it simplify interpretation or enhance model performance?\

4.  What are the limitations of this coding method? Are there situations where this method may lead to misleading results or interpretations?\

### Applying Simple Coding

```{r}
# assign simple coding
#creating the contrast matrix manually by modifying the dummy coding scheme
c <- contr.treatment(4)
my.coding <- matrix(rep(1/4, 12), ncol=3)
my.simple <- c-my.coding
my.simple


contrasts(hsb2$race.f) <- my.simple
summary(lm(write ~ race.f, hsb2))


# build and summarize your regression
simple <- lm(write ~ race.f, data = hsb2)
summary(simple)

?contr.treatment
```

Summarize these findings. What is your reference group, and how do the other groups compare with regards to their writing scores? What does the intercept value represent?\
\
(write here)\

## Deviation Coding (a.k.a. Effect Coding)

(Alii, Bennett, LaVelle)

1.  What comparisons are made with this coding scheme? How does this coding scheme affect the coefficients in regression models? (What do the coefficients represnt? What is the reference level?)\

2.  When is this coding scheme preferred? What types of data or analysis scenarios make this coding method particularly useful?\

3.  What are the advantages of this coding method? How does it simplify interpretation or enhance model performance?\

4.  What are the limitations of this coding method? Are there situations where this method may lead to misleading results or interpretations?

### Applying Deviation Coding

```{r}
# assign deviation coding
contr.sum(4)


#assigning the deviation contrasts to race.f
contrasts(hsb2$race.f) = contr.sum(4)
#the regression
summary(lm(write ~ race.f, hsb2))

hsb2$race.f <- relevel(hsb2$race.f, ref = "Asian")


# build and summarize your regression
dev <- lm(write ~ race.f, data = hsb2)
summary(dev)

# Get the mean of the fourth group (e.g., "Caucasian")
mean_caucasian <- mean(subset(hsb2, race.f == "Caucasian")$write)

# Compare it to the grand mean (intercept value)
grand_mean <- coef(lm(write ~ race.f, data = hsb2))["(Intercept)"]


means_tapply <- tapply(hsb2$write, hsb2$race.f, mean)
print(means_tapply)

m_o_m <- (46.45833+58+48.2+54.05517)/4
g_m <- mean(hsb2$write)

summary(hsb2$race.f)
```

Summarize these findings. Unlike dummy and simple coding, there is no traditional reference level in deviation coding. In this context, how do the groups compare regarding their writing scores? What does the intercept value represent?\
\
(write here)\

## Helmert Coding

(Berylin)

1.  What comparisons are made with this coding scheme? How does this coding scheme affect the coefficients in regression models? (What do the coefficients represnt? What is the reference level?)

\- Helmert coding checks if one specific group stands out from the crowd of everything that follows it.

-   **Round 1:** Group 1 vs. Average of (2, 3, and 4)

-   **Round 2:** Group 2 vs. Average of (3 and 4)

-   **Round 3:** Group 3 vs. Group 4

    -   Instead of comparing everyone to a "Control Group," you compare each group to the **average of everyone still waiting in line.**

\- In a regression model, the coefficient is just a number that tells you the **size of the gap** between one group and the groups that follow it.

-   **A Positive Coefficient:** Means this group performed **better** than the average of the groups coming after it.

-   **A Negative Coefficient:** Means this group performed **worse** than the average of the groups coming after it.

-   **Zero:** Means this group is **exactly the same** as the average of the groups that follow.

> **Simple Template:** "How much does Group X stand out from the average of the groups that are left?"

-   **There is no single reference level.**

\- In standard "Dummy Coding," you pick one group (like a Placebo) and compare everyone to it. In Helmert coding, the **reference point moves.**

-   For the first comparison, the reference is the average of 2, 3, and 4.

-   For the second comparison, the reference shifts to the average of 3 and 4.

-   The **last group (Group 4)** never gets its own coefficient; it only exists to serve as the "final comparison" for Group 3.

2.  When is this coding scheme preferred? What types of data or analysis scenarios make this coding method particularly useful?

\- Use it for Sequential Progression: At what specific step in this sequence does the result significantly change?

-   Unlike other methods that compare everything to a single "Control" group, Helmert coding is used when you want to see if a level stands out from the "future" levels that follow it. It is the best choice when the **order** of your categories is the most important part of your theory.

\- To use Helmert coding effectively, your data must be **Ordinal**. This means the categories have a natural, logical rank or direction.

-   **Progressive Treatments:** (None $\rightarrow$ Low Dose $\rightarrow$ Medium Dose $\rightarrow$ High Dose). You use it to see if "Low Dose" differs from the average of the stronger doses.

-   **Time-Based Phases:** (Baseline $\rightarrow$ Week 1 $\rightarrow$ Week 4 $\rightarrow$ Week 12). You use it to see if the initial results at Week 1 are different from the long-term average of the later weeks.

-   **Developmental Stages:** (Infancy $\rightarrow$ Childhood $\rightarrow$ Adolescence). You use it to pinpoint exactly which stage of growth creates a statistical "jump" compared to the older stages.

\- Helmert coding is particularly useful in these two scenarios:

-   **Finding Thresholds:** You want to know the "minimum viable" level. For example, "Is a Bachelor’s degree enough to stand out from the average of all higher graduate degrees?"

-   **Testing Cumulative Effects:** You want to see if moving one step further in a process adds value compared to the combined weight of all remaining steps.

3.  What are the advantages of this coding method? How does it simplify interpretation or enhance model performance?

    \- Helmert coding shows you a step-by-step story instead of just saying "these groups are different."

    -   Finds where the change happens: Shows you exactly which step in the ladder matters (like "Bachelor's to Master's makes a difference, but Master's to PhD doesn't")

    -   Answers "Is the next step worth it?": Tells you if moving up one level actually helps compared to the levels above it

    -   Keeps it simple: Instead of comparing every group to every other group, you just follow the steps up the ladder\
        \

4.  What are the limitations of this coding method? Are there situations where this method may lead to misleading results or interpretations?

**-** Helmert coding only works when categories have a natural ranking, using it incorrectly produces meaningless results.

-   Requires ordered categories: Doesn't work for unranked groups (like race, country, or brands) because it forces a hierarchy where none exists

-   Ordering changes everything: Rearranging the sequence gives completely different conclusions, so you must have a solid reason for your order

-   Complex to communicate: Explaining "compared to the average of later groups" is harder than "compared to the control group"

-   Not for all comparisons: Can't answer questions like "Is Treatment A better than Treatment B directly?"—it compares to averages instead

-   Assumes step-by-step progression: Breaks down when the pattern isn't a simple ladder (like when middle levels perform better than both low and high)\

### Applying Helmert Coding

```{r}
# assign Helmert coding
#helmert for factor variable with 4 levels
my.helmert = matrix(c(3/4, -1/4, -1/4, -1/4, 0, 2/3, -1/3, -1/3, 0, 0, 1/
	2, -1/2), ncol = 3)
my.helmert
      

#assigning the new helmert coding to race.f
contrasts(hsb2$race.f) = my.helmert
contrasts(hsb2$race.f)

hsb2$race.f <- relevel(hsb2$race.f, ref = "Caucasian")
contrasts(hsb2$race.f) = my.helmert

# build and summarize your regression
helmert <- lm(write ~ race.f, data = hsb2)
summary(helmert)
```

Summarize these findings. In Helmert coding, each level is compared to the mean of subsequent levels, rather than having a traditional reference group. In this context, how do the groups compare regarding their writing scores? What does the intercept value represent?\
\
Answer:

Caucasian students outperform other groups collectively, Hispanic students underperform compared to Asian and African-American students, and Asian students significantly outscore African-American students, with an overall average writing score of 51.68 points across all groups.

-   **"Column 1 → race.f1 = +3.17: Caucasian beats the field by 3.17"**

-   **"Column 2 → race.f2 = -6.64: Hispanic loses to Asian/African-Am by 6.64"**

-   **"Column 3 → race.f3 = +9.80: Asian beats African-American by 9.80"**

    For each comparison, one group challenges several other groups. The positive weight equals the number of groups being challenged divided by the total number in that matchup. Each challenged group gets the negative version of one divided by the total number in the matchup.

-   **1 vs. 3 groups:** Positive = 3/4, Negatives = -1/4 each

-   **1 vs. 2 groups:** Positive = 2/3, Negatives = -1/3 each

-   **1 vs. 1 group:** Positive = 1/2, Negative = -1/2

[The code is re-arranging the groups so that instead of comparing everyone to one "baseline" group, you are comparing them in a descending "knockout" style to see where the biggest performance gaps are.]{.underline}

## For next class: Releveling a predictor

We will do this together as a class. You don't need to work ahead, but if you wanted to explore releveling, here you go.

Releveling refers to the process of changing the reference level of a factor or categorical variable. This allows us to examine how different groups compare to a new reference category. By selecting a different reference level, we can gain new insights into the effects of categorical predictors in our analyses.

Releveling is applicable to Dummy, Simple, and Helmert coding, but NOT deviation coding. Why?

### To relevel:

You can use the below to define a reference level, or you can assign new values in your contrasts matrix.

```{r}
hsb2$race.f <- relevel(hsb2$race.f, ref = "Asian")
```
